%% ============================================================
%% Foundations of Sensitivity Analysis: Theory, Computation, and Applications
%% Leon M. Arriola and James M. Hyman
%%
%% Master book file — compile with pdflatex (two passes)
%% ============================================================

\documentclass[12pt]{book}

%% ---- Shared preamble (packages, macros, environments) ----
\input{preamble}

%% ---- Book metadata ----
\title{%
  \textbf{Foundations of Sensitivity Analysis}\\[6pt]
  \large Theory, Computation, and Applications
}
\author{Leon M. Arriola \and James M. Hyman}
\date{}

%% ============================================================
\begin{document}
%% ============================================================

\frontmatter

%% ---- Title page ----
\maketitle

%% ---- Copyright / colophon ----
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
\small
\copyright\ 2026 Leon M. Arriola and James M. Hyman.\\
All rights reserved.
\end{center}
\vspace*{\fill}
\clearpage

%% ---- AMS Classification and Keywords ----
\thispagestyle{empty}
\vspace*{\fill}
\begin{flushleft}\small
\textbf{AMS subject classifications.}
65-01,
34A55,
62-07,
92D30,
49K15,
65C05

\medskip
\textbf{Keywords.}
sensitivity analysis,
adjoint method,
forward sensitivity equations,
Sobol indices,
global sensitivity analysis,
uncertainty quantification,
mathematical modeling,
epidemic models,
automatic differentiation
\end{flushleft}
\vspace*{\fill}
\clearpage

%% ---- Table of contents ----
\tableofcontents
\clearpage

%% ---- Preface ----
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book develops sensitivity analysis (SA) as a mathematical discipline,
not merely as a collection of computational recipes.
SA asks the foundational question of quantitative modeling:
which parameters most strongly determine the model's predictions?

\medskip

Three ideas run through every chapter:
\begin{enumerate}[nosep]
  \item The sensitivity index is a normalized derivative.
  \item The adjoint is the transpose, in whatever space you work.
  \item Local SA is a linearization, with all the limitations that implies.
\end{enumerate}

Students who understand these three ideas (not just their definitions
but their implications and connections) will be equipped to design,
implement, and critique SA studies in any application domain.

\medskip

The SIR epidemic model serves as the running example throughout.
Every method is demonstrated on this one model, so students can
track increasing mathematical sophistication against a fixed biological
context.
All computational examples use MATLAB; each chapter includes a
structured laboratory and a reusable code template.
All code is freely available in the companion repository:
\begin{center}
  \texttt{https://github.com/mhyman/arriola2026sa}
\end{center}
The repository contains the complete \texttt{src/} library of
production-quality MATLAB functions (one per major algorithm),
three test suites with 48 automated checks, and this instructor's
solutions manual.

\medskip

The adjoint method (Chapters~5--7) is the
conceptual centerpiece of the book. It is the same idea encountered
three times in three different mathematical spaces: the matrix transpose
in linear algebra, the reversed computation graph in algorithmic
differentiation, and a backward ODE in dynamical systems.
Chapter~8, \emph{Caveat Emptor}, asks when sensitivity analysis can
mislead, and is the chapter students most frequently cite as the
one that changed how they think about mathematical models.

\medskip

\noindent\textbf{Relationship to existing literature.}
Cacuci (1981) established the adjoint formalism for nonlinear systems;
Giles and Pierce (2000) developed it for aeronautical design;
Cao et al.\ (2002) gave the computational treatment for ODE systems;
and Griewank and Walther (2008) treat the computation-graph view
comprehensively. The present book differs from each of these in its
pedagogical aim: it is the only undergraduate text that introduces all
three forms of the adjoint (matrix, graph, and ODE) as instances
of a single unifying idea, and that teaches students when \emph{not}
to trust the results they compute.

\medskip

\noindent\textbf{For instructors.}
The book covers a one-semester undergraduate course.
Students need calculus, linear algebra, and ODEs; no measure theory or
functional analysis is assumed.
A minimal one-semester course can omit the computational AD material
in Chapter~7 and all appendices without loss of continuity.
Chapters~1--4 and~8--9 form a self-contained global-SA course
for students who will not use the adjoint in their research.
Exercises tagged \textsc{[Bloom~L5]} and \textsc{[L6]} are designed for
semester projects and take-home examinations; those tagged \textsc{[L2]}--\textsc{[L4]}
are suitable for weekly problem sets.

\medskip

\medskip

\noindent\textit{Leon M. Arriola, University of Wisconsin-Whitewater}\\
\noindent\textit{James M. Hyman, Tulane University}

\medskip

\noindent\textbf{Acknowledgments.}
We thank the students in our sensitivity analysis courses at the
University of Wisconsin-Whitewater, Tulane University, and the
Mathematical and Theoretical Biology Institute (MTBI) at Arizona
State University, whose questions and confusion shaped every page
of this book.

The MATLAB code library in the companion GitHub repository
(\texttt{https://github.com/mhyman/arriola2026sa}) was
developed with support from the Tulane University Department of
Mathematics.

\medskip

\noindent\textbf{Dedication.}

\medskip
\begin{center}
\textit{To our students, past and future:} \\[4pt]
\textit{May you always ask which parameters matter} \\
\textit{before you run the model.}
\end{center}

%% ============================================================
\mainmatter
%% ============================================================

%% ---- Part I: The Sensitivity Question ----
\part{The Sensitivity Question}

\input{ch01_body}
\input{ch02_body}

%% ---- Part II: Computing SA in Practice ----
\part{Computing Sensitivity in Practice}

\input{ch03_body}

%% ---- Part III: Analytic Forward SA ----
\part{Analytic Forward Sensitivity Analysis}

\input{ch04_body}

%% ---- Part IV: The Adjoint Approach ----
\part{The Adjoint Approach}

\input{ch05_body}
\input{ch06_body}
\input{ch07_body}

%% ---- Part V: Limitations and Extensions ----
\part{Limitations and Extensions}

\input{ch08_body}
\input{ch09_body}

%% ============================================================
\appendix
%% ============================================================

\part*{Appendices}
\addcontentsline{toc}{part}{Appendices}

\input{appendix_a}
\input{appendix_b}
\input{appendix_c}
\input{appendix_d}
\input{appendix_e}

%% ============================================================
\backmatter
%% ============================================================

%% ---- Bibliography ----
\bibliographystyle{siam}
\bibliography{references}

%% ---- Index ----
\printindex

\end{document}
